Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-19 paper/paper.tex
paper: refine chapters 1-4

In this paper, we focus on PDP and ALE and refer to them when speaking of
feature effects.

@@

Test functions play a crucial role in research, e.g.\ for evaluating different
methodological approaches, or when conducting simulation studies. In the
following, we synthesize commonly used test functions across various domains,
providing structured guidance for researchers --- particularly those in
Interpretable ML --- in selecting appropriate test functions for their
simulation studies. By examining test functions from different fields and their
purposes of application, we aim to facilitate experimental design decisions
rather than giving an exhaustive overview.

\paragraph{Test Functions in Optimization.}
The field of optimization, where test functions are essential to enable the
assessment and comparison of optimization algorithms, has established a rich
foundation of test functions. A fundamental approach involves using simple
mathematical expressions like the sphere function~\cite{more_testing_1981}.
These basic functions are often combined with more complex ones like Branin or
Rosenbrock to create comprehensive test suites that incorporate important
properties such as nonlinearity, non-separability, and
scalability~\cite{whitley_building_1998}. A notable framework in this domain is
the Comparing Continuous Optimizer (COCO) platform with its Black Box
Optimization Benchmark (BBOB), offering a structured approach to testing
continuous optimization algorithms through artificial test
functions~\cite{hansen_coco_2016}. These classical test function suites are
well-established in optimization and may also serve as a basis for
Interpretable ML researchers. However, the ability of these artifical test
functions to represent complex real-world behavior is often
limited~\cite{zaefferer_simulation-based_2017}.

@@

Physics-derived functions offer a compelling source of real-world test cases,
with the \textit{Feynman Symbolic Regression Database
    (FSReD)}~\cite{udrescu_ai_2020} being a prominent example. FSReD comprises 100
physics equations from the seminal \textit{Feynman's Lectures on Physics
    (34--36)}, supplemented by 20 more challenging equations from other seminal
physics texts. These equations span diverse physics domains and involve a
varying number of variables and various elementary functions such as arithmetic
operations, trigonometric functions, and exponentials. Tabular datasets are
generated through random sampling from defined value ranges.

Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
of the original FSReD. They introduce a three-tiered categorization of problems
(easy, medium, hard) based on their complexity, incorporate dummy variables to
simulate irrelevant features, and implement more realistic sampling ranges and
strategies. Detailed specifications for all formulas, including their sampling
parameters, are available in their work.

While initially developed for symbolic regression tasks, many Feynman equations
may serve as suitable test functions for simulation studies in Interpretable
ML. Their basis in physical principles provides real-world relevance, though  % chktex 13
researchers should carefully select equations that align with their specific
analytical objectives and complexity requirements.

\paragraph{Test Functions for Interpretable Machine Learning.}
The Interpretable ML field itself has developed several specialized test
functions designed to evaluate specific aspects of interpretability methods.

Goldstein et al.~\cite{goldstein_peeking_2015} used several simple test
functions to demonstrate the behavior of Individual Conditional Expectation
(ICE) curves. These include a simple additive function to demonstrate the
absence of interactions, simple interactions to reveal heterogeneity that might
be obscured by averaging procedures such as PDPs, and a specially designed
function with an empty quadrant for assessing extrapolation behavior.

Similarly, Liu et al.~\cite{liu_model_2018} focus on simple functions before
progressing to more complex ones. They begin with basic two-variable scenarios
--- using additive functions, interaction functions, and combinations thereof
--- and examine these under both independent and correlated feature conditions
to compare various feature effect methods. The advantage of these simple test
functions is that solutions (e.g., feature effects) can also be computed
analytically, and that they allow for deeper and more fine-grained analysis of
individual aspects.

A more complex test function suite was proposed by
Tsang~et~al.~\cite{tsang_detecting_2017}, specifically designed to evaluate the
detection of variable interactions. Their functions incorporate various types
of interactions with different orders, strengths, non-linearities, and
overlaps. While this makes them particularly valuable for interaction
detection, they are also useful for evaluating other interpretability methods
in scenarios with complex interactions.

@@

When choosing test functions for simulation studies in Interpretable ML,
researchers should consider several criteria, including the specific aspects of
interpretability being evaluated, the desired complexity level and number of
variables, the presence of specific challenges such as correlation between
features or interactions, the need for analytical solutions for validation, as
well as the relevance to real-world applications in the domain of interest.

\section{Background}\label{sec:background}

@@

The \textit{Partial Dependence Plot (PDP)} by
Friedman~\cite{friedman_greedy_2001} describes the marginal effect of one or
two features on the prediction of a model $\hat f$. For a feature set $X_S$
(with $S \subseteq \{1,\ldots,p\}$, $|S| = 1$ or $|S| = 2$), the PDP is defined
as

\begin{equation}
    PDP_{\hat f, S} = \mathbb{E}_{X_C}[\hat{f}(x_S, X_C)] = \int f(x_S, x_C)d\mathbb{P}(x_C),
\end{equation}

@@

\noindent Here, $x_C^{(i)}$ are the actual complement feature values from the dataset of $n$ instances.
To plot this function, a grid of $G$ grid points
$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.  % chktex 3
Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.\\

\noindent The \textit{Accumulated Local Effects (ALE)} plot is an alternative to the PDP that
solves the extrapolation issue~\cite{apley_visualizing_2020}. Using the notation above,
for $|S|=1$, the ALE plot is defined as

@@

\noindent where $\hat{f}^S(x_S, x_C) = \frac{\partial \hat{f}(x_S, x_C)}{\partial x_S}$.
The constant is chosen so that $\widehat{ALE}_{\hat f,S}(X_S)$ is centered with a mean of $0$
w.r.t.\ the marginal distribution of $X_S$. The uncentered ALE can be estimated by

@@

\noindent Here, $\{N_S(k) = (z_{k-1,S}, z_{k,S}]\}_{k=1}^{K}$ partitions the samples  % chktex 3 chktex 9
$\{x^{(i)}_S\}_{i=1}^n$ into $K$ intervals or neighborhoods $N_S(k)$. $n_S(k)$ denotes  % chktex 3
the number of observations in the $k$th interval $N_S(k)$, $k_S(x)$ represents the index
of the interval to which a particular value $x$ of feature $x_S$ belongs.
The uncentered ALE is centered by

@@

\noindent to have a mean effect of $0$. For the grid that defines the intervals,
the quantiles of the empirical distribution of $\{x^{(i)}_S\}_{i=1}^n$ can be  % chktex 3
used~\cite{apley_visualizing_2020}.\\

@@

To quantify the error of a computed feature effect, a ``ground truth'' needs to
be defined first. We follow the approach of Molnar et
al.~\cite{molnar_relating_2023} and define ground truth versions of PDP and ALE
directly on the data generating process (DGP) by applying PDP and ALE to the
underlying ground truth function $f$ (instead of model $\hat f$).

For PDP, we can directly use the definition of Molnar et
al.~\cite{molnar_relating_2023}:

@@

As stated by Molnar et al.~\cite{molnar_relating_2023}, their results also
apply to conditional variants of the PDP such as ALE. We now make this  % chktex 13
definition explicit:

@@

%With these definitions, $\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$
%can now be treated as statistical estimators of the ground truth feature effects
%(cf.~\cite{molnar_relating_2023}). 
Note that different ground truth effects may also be defined, and our choices
come with certain implications and limitations, such as omitting the
\textit{aggregation bias}\footnote{for details,
    see~\cite{herbinger_repid_2022}}~\cite{mehrabi_survey_2021}.

With more complex ground truth functions $f$, it may become increasingly
difficult to derive the ground truth feature effects analytically, especially
for ALE. In these cases, we therefore propose to also estimate those effects by  % chktex 13
Monte Carlo integration, yielding $\widehat{PDP}_{f,S}(x_S)$ and
$\widehat{ALE}_{f,S}(x_S)$ (obtained by plugging in $f$ instead of $\hat f$
into the estimators in equations (\ref{eq:pdp-estimate}) and
(\ref{eq:ale-estimate-uncentered}) / (\ref{eq:ale-estimate-centered})).\\

\noindent Summarizing, we have now defined four quantities per feature effect:
$PDP_{f,S}$ $\widehat{PDP}_{f,S}$, $PDP_{\hat f,S}$, and $\widehat{PDP}_{\hat f,S}$
(analogue for ALE). We can now define different errors between each of these
quantities. In this paper, we focus on the MSE, as it can be decomposed into
bias and variance (see e.g.~\cite{geman_neural_1992}).
Taking, for example, $PDP_{f,S}$ as ground truth, we can define the MSE of
$PDP_{\hat f,S}$ at a point $x_S$ as follows~\cite{molnar_relating_2023}:

@@

Since we usually cannot determine $PDP_{\hat f,S}$, we need to estimate it by
Monte Carlo integration, yielding $\widehat{PDP}_{\hat f,S}(x_S)$. This,
however, introduces an additional variance term (we use the random variable
$X_{mc}$ for the Monte Carlo samples (e.g., training or validation data)):

\begin{equation}
    \begin{split}
        \text{MSE}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S})
        = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f}, S}(x_S)])}^2}_{Bias^2} \\
        + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance} + \underbrace{\mathbb{E}_F\text{Var}_{X_{mc}}[\widehat{PDP}_{\hat{f},S}(x_S)]}_{MC-Variance}
    \end{split}
\end{equation}

@@

\noindent We see that the variance due to MC integration also depends on the model
distribution $F$. Similarly, one could use the estimate $\widehat{PDP}_{f,S}$
as groundtruth, introducing an additional variance term $Var_{X_{mc2}}$when
estimated on different MC sample (proof in
\textsc{Appendix~\ref{app:proof-mc-variance}}).
An overview of all error terms in this chain can be found in
\textbf{Fig.\@~\ref{fig:error-graph}}.\\

\begin{figure}[t]
    \includegraphics[width=\textwidth]{img/error_graph.pdf}
    \caption{Error chain in feature effect estimation
        (modified, original version can be found in~\cite{molnar_relating_2023})}\label{fig:error-graph}
\end{figure}

\noindent $\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets sampled independently from the DGP. We propose the following estimators:  % chktex 13
\begin{equation}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
    \label{eq:mse-estimator}
\end{equation}
\begin{equation}
    \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
    \label{eq:bias-estimator}
\end{equation}
\begin{equation}
    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
    \frac{1}{M-1}\sum_{m=1}^M{\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)}^2
    \label{eq:variance-estimator}
\end{equation}

\noindent Note that these are similar to the approach in~\cite{molnar_relating_2023},
but we do not specify which data points to use for MC integration. The
variance captures both the variance in the model fits and the variance due to
MC integration. To estimate the MC variance, we propose the following estimators:
\begin{equation}
    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K{(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))}^2
    \label{eq:mc-variance-estimator-groundtruth}
\end{equation}

\noindent and
\begin{equation}
    \begin{split}
        \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
        {\left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)}^2
    \end{split}
    \label{eq:mc-variance-estimator-model}
\end{equation}

\noindent For more convenient analysis of the errors, one could also
aggregate them over the marginal distribution of $X_S$ (e.g.,
estimated by averaging over the grid points if chosen appropriately)
to obtain a single error measure per feature effect.

While our definitions are based on the PDP, they can be directly applied to the
ALE as well.

@@

To address the research question of which data to choose to estimate feature
effects, we conduct a comprehensive simulation study. For different models,
datasets, and dataset sizes, we estimate the feature effects PDP and ALE on
training data, validation data, and in a cross-validated manner. We then
compute the feature effect error as MSE with respect to the ground truth
feature effects and decompose it into bias and variance components. By
comparing these error terms across the different estimation strategies
(training, validation, cross-validation), we aim to provide nuanced empirical
insights into the implications of the data choice on feature effect estimation.
For deeper analysis, we conduct two ablation studies: one decomposing the
feature effect variance into model variance and Monte Carlo variance, and
another examining the impact of the dataset size on the Monte Carlo variance.
In the following, we describe our experimental set-up for the main simulation
study (\textsc{Section~\ref{sec:experiment-main}}) and the ablation studies
(\textsc{Section~\ref{sec:experiment-ablation}}).

@@

\begin{itemize}[label=--]
    \item \textbf{SimpleNormalCorrelated} consists of four
          standard-normally distributed features, where the first two
          features exhibit strong correlation ($\rho = 0.9$) while the others are
          independent dummy variables. The target variable is given by this simple
          formula:
          \begin{equation}
              f_1(\boldsymbol{x}) = x_1 + \frac{x_2^2}{2} + x_1 x_2
          \end{equation}
          This test function is inspired by~\cite{liu_model_2018} and aims
          to focus on the impact of correlation and interactions.
    \item \textbf{Friedman1} implements the classical Friedman1 benchmark
          function~\cite{breiman_bagging_1996,friedman_multivariate_1991}
          with seven uniformly distributed features between 0 and 1, all mutually
          independent:
          \begin{equation}
              f_2(\boldsymbol{x}) = 10 \sin(\pi x_1 x_2) + 20{(x_3 - \frac{1}{2})}^2 + 10 x_4 + 5 x_5
          \end{equation}
          This dataset includes a mix of linear and different non-linear effects with
          interactions.
    \item \textbf{Feynman I.29.16} is based on the Feynman equation I.29.16
          describing wave interference, comprising six independent features. Using
          the refined sampling strategies from Matsubara et al.~\cite{matsubara_rethinking_2024},
          two log-uniformly distributed variables on $[0.1, 10]$, two angles uniformly
          distributed over $[0, 2\pi]$, and two uniformly distributed dummy variables on [0,1]:
          \begin{equation}
              f_3(\boldsymbol{x}) = \sqrt{x_1^2 + x_2^2 + 2 x_1 x_2 \cos(\theta_1 - \theta_2)}
          \end{equation}
          This dataset is designed to reflect a physics-based relationship for more real-world relevance.

@@

\noindent To generate the datasets, a standard normally distributed noise term $\epsilon$ is added to each function,
scaled by a signal-to-noise ratio of five\footnote{To determine the factor by which the noise is multiplied,
    the standard deviation of the signal is computed over 100'000 randomly drawn samples of $y$ and divided by
    the signal-to-noise ratio of five.}. For each test function, we consider two dataset sizes:
$(n_{train}=1000, n_{val}=250)$ and $(n_{train}=8000, n_{val}=2000)$.

@@

\paragraph{Feature effect estimation.}
For the trained models, we estimate the feature effects
$\widehat{PDP}_{\hat{f},S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
Monte Carlo integration, we use for each the full training and validation set
as well as a cross-validation strategy. In the latter case, we use a 5-fold
cross-validation on the combined samples of training and validation set, in
each fold fit the model on four folds, and estimate the feature effects on the
remaining fold. We then average the feature effects over all folds. We also
estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
$\widehat{ALE}_{f,S}$ since the complexity of the Feynman equation makes
analytical computation infeasible. For the Monte Carlo integration, we use
10000 samples additionally sampled from the DGP, which introduces only a
negligible additional variance term as we show in (\dots) We use 100 grid
points, defined by the quantiles of the theoretical distribution of $X_S$ to
enable a comparison across feature effects. Additionally, we center the curves
after estimation to have a mean effect of $0$ w.r.t.\ the grid points, and we
omit the first and the last grid point after estimation to avoid boundary
effects especially occuring in ALE plots.

>>>>

In the following, we focus on PDP and ALE and refer to them when speaking of
feature effects.

@@

Test functions play a crucial role in research, e.g.\ for evaluating different
methodological approaches, or when conducting simulation studies. In the
following, we synthesize commonly used test functions across various domains,
providing structured guidance for researchers, particularly those in
Interpretable ML, in selecting appropriate test functions to facilitate
designing simulation studies.

\paragraph{Test Functions in Optimization.}
The field of optimization, where test functions are essential to enable the
assessment and comparison of optimization algorithms, has established a rich
foundation of test functions. A fundamental approach involves using simple
mathematical expressions like the sphere function~\cite{more_testing_1981}.
These basic functions are often combined with more complex ones like Branin or
Rosenbrock to create comprehensive test suites that incorporate important
properties such as nonlinearity, non-separability, and
scalability~\cite{whitley_building_1998}. A notable framework in this domain is
the Comparing Continuous Optimizer (COCO) platform with its Black Box
Optimization Benchmark (BBOB), offering a structured approach to testing
continuous optimization algorithms through artificial test
functions~\cite{hansen_coco_2016}. While being
well-established in optimization, they may also serve as a basis for
Interpretable ML researchers. However, the ability of these artificial test
functions to represent complex real-world behavior is
limited~\cite{zaefferer_simulation-based_2017}.

@@

Physics-derived functions offer a compelling source of real-world test cases,
with the \textit{Feynman Symbolic Regression Database
    (FSReD)}~\cite{udrescu_ai_2020} as a prominent example. FSReD comprises 100
physics equations from the seminal \textit{Feynman's Lectures on Physics
    (34--36)}, supplemented by 20 more challenging equations from other seminal
physics texts. These equations span diverse physics domains and involve a
varying number of variables and various elementary functions. Tabular datasets are
generated through random sampling from defined value ranges.

Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
of the original FSReD by introducing a three-tiered categorization of problems
based on their complexity, incorporating dummy variables to
simulate irrelevant features, and implementing more realistic sampling ranges and
strategies. Detailed specifications for all formulas, including their sampling
parameters, are available in their work.

While originally designed for symbolic regression tasks, many Feynman equations
serve as suitable test functions for simulation studies in Interpretable
ML. Their basis in physical principles provides real-world relevance, though  % chktex 13
researchers should carefully select equations that align with their specific
analytical objectives and complexity requirements.

\paragraph{Test Functions for Interpretable Machine Learning.}
The Interpretable ML field itself has developed several specialized test
functions designed to evaluate specific aspects of interpretability methods.
Goldstein et al.~\cite{goldstein_peeking_2015} used several simple test
functions to demonstrate the behavior of Individual Conditional Expectation
(ICE) curves. These include a simple additive function to demonstrate the
absence of interactions, simple interactions to reveal heterogeneity that might
be obscured by PDPs, and a specially designed
function with an empty quadrant for assessing extrapolation behavior.

Similarly, Liu et al.~\cite{liu_model_2018} focused on simple functions before
increasing complexity. They started with basic two-variable scenarios
--- additive functions, interaction functions, and combinations thereof
--- and examined these under both independent and correlated features,
comparing various feature effect methods. Advantages of these simple test
functions are that solutions (e.g., feature effects) can be computed
analytically, and that they allow for more fine-grained analyses of
individual aspects.

A more complex test function suite was proposed by
Tsang~et~al.~\cite{tsang_detecting_2017}, specifically designed to evaluate the
detection of variable interactions. Their functions incorporate various types
of interactions with different orders, strengths, non-linearities, and
overlaps. While being particularly valuable for interaction 
detection, they are also useful for evaluating other interpretability methods
in scenarios with complex interactions.

@@

When choosing test functions for simulation studies in Interpretable ML,
researchers should consider several criteria, including the specific aspects of
interpretability being evaluated, the desired complexity level and number of
variables, the presence of specific challenges such as correlation between
features or interactions, the need for analytical solutions for validation, as
well as the relevance to real-world applications.

\section{Theoretical Framework: Feature Effects and Error Decomposition}\label{sec:background}

@@

The \textit{Partial Dependence Plot (PDP)}~\cite{friedman_greedy_2001} describes the marginal effect of one or
two features on the prediction of a model $\hat f$. For a feature set $X_S$
(with $S \subseteq \{1,\ldots,p\}$, $|S| = 1$ or $|S| = 2$), the PDP is defined
as

\begin{equation}
    PDP_{\hat f, S} = \mathbb{E}_{X_C}[\hat{f}(x_S, X_C)] = \int \hat f(x_S, x_C)d\mathbb{P}(x_C),
\end{equation}

@@

\noindent Here, $x_C^{(i)}$ are the actual complement feature values from the dataset of $n$ instances.
To plot the function, a grid of $G$ grid points
$\{(x_S^{(g)}, \widehat{PDP}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.  % chktex 3
Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.\\

\newpage
\noindent The \textit{Accumulated Local Effects (ALE)} plot~\cite{apley_visualizing_2020}
for $|S|=1$ is defined as

@@

\noindent where $\hat{f}^S(x_S, x_C) = \frac{\partial \hat{f}(x_S, x_C)}{\partial x_S}$.
The constant is chosen so that $ALE_{\hat f,S}(X_S)$ is centered with a mean of $0$
w.r.t.\ the marginal distribution of $X_S$. The uncentered ALE can be estimated by

@@

\noindent Here, $\{N_S(k) = (z_{k-1,S}, z_{k,S}]\}_{k=1}^{K}$ partitions the samples  % chktex 3 chktex 9
$\{x^{(i)}_S\}_{i=1}^n$ into $K$ intervals or neighborhoods $N_S(k)$. $n_S(k)$ denotes  % chktex 3
the number of observations in the $k$th interval, $k_S(x)$ the index
of the interval to which a particular value $x$ of feature $S$ belongs.
The uncentered ALE is centered by

@@

\noindent to have a mean effect of $0$. For the grid that defines the intervals,
the quantiles of the empirical distribution of $\{x^{(i)}_S\}_{i=1}^n$ can be  % chktex 3
used~\cite{apley_visualizing_2020}.

@@

\paragraph{Feature effect ground truth.} To quantify the error of a computed
feature effect, a ``ground truth'' needs to
be defined first. We follow the approach of Molnar et
al.~\cite{molnar_relating_2023} and define ground truth versions of PDP and ALE
directly on the data generating process (DGP) by applying PDP and ALE to the
underlying ground truth function $f$ (instead of model $\hat f$).
For PDP, we can directly use the definition of Molnar et
al.~\cite{molnar_relating_2023}:

@@

\noindent As stated by Molnar et al.~\cite{molnar_relating_2023}, their results also
apply to conditional variants of the PDP such as ALE. We now make this  % chktex 13
definition explicit.

\newpage

@@

%With these definitions, $\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$
%can now be treated as statistical estimators of the ground truth feature effects
%(cf.~\cite{molnar_relating_2023}). 
\noindent Note that different ground truth feature effects can also be defined, and our choices
come with certain implications and limitations, such as omitting the
\textit{aggregation bias}~\cite{herbinger_repid_2022,mehrabi_survey_2021}.

With more complex ground truth functions $f$, it becomes often infeasible to
derive the ground truth feature effects analytically. For these cases, we
therefore propose to also estimate those effects by  % chktex 13
Monte Carlo integration, yielding $\widehat{PDP}_{f,S}(x_S)$ and
$\widehat{ALE}_{f,S}(x_S)$ (obtained by plugging $f$ instead of $\hat f$
into the estimators in equations (\ref{eq:pdp-estimate}) and
(\ref{eq:ale-estimate-uncentered}) / (\ref{eq:ale-estimate-centered})).

\paragraph{Feature effect error decomposition.} Summarizing, we have now defined four quantities per feature effect:
$PDP_{f,S}$ $\widehat{PDP}_{f,S}$, $PDP_{\hat f,S}$, and $\widehat{PDP}_{\hat f,S}$
(analogously for ALE). Between these quantities, different error metrics can because
defined. In this paper, we focus on MSE as it can be decomposed into
bias and variance~\cite{geman_neural_1992}.
When taking $PDP_{f,S}$ as ground truth, the MSE of $PDP_{\hat f,S}$ at
point $x_S$ can be formulated as follows~\cite{molnar_relating_2023}:

@@

Since we usually cannot determine $PDP_{\hat f,S}$, we need to estimate it by
Monte Carlo integration, which introduces an additional variance term (we use the random variable
$X$ for the Monte Carlo samples):

\begin{equation}
    \begin{split}
        \text{MSE}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S})
        = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f}, S}(x_S)])}^2}_{Bias^2} \\
        + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance} + \underbrace{\mathbb{E}_F\text{Var}_{X}[\widehat{PDP}_{\hat{f},S}(x_S)]}_{MC-Variance}
    \end{split}
    \label{eq:mse-decomposition}
\end{equation}

@@

\noindent We see that the variance due to MC integration also depends on the model
distribution $F$. Similarly, one could use the estimate $\widehat{PDP}_{f,S}$
as ground truth, introducing an additional variance term when
estimated with Monte Carlo samples (proof in
\textsc{Appendix~\ref{app:proof-mc-variance}}).
An overview of all error terms is given in
\textbf{Fig.\@~\ref{fig:error-graph}}.

\begin{figure}[t]
    \includegraphics[width=\textwidth]{img/error_graph.pdf}
    \caption{Error chain in feature effect estimation. The deviation of the estimated
        feature effect from the ground truth can be decomposed into bias, model variance,
        and Monte Carlo variance. If the ground truth is estimated by Monte Carlo
        integration, an additional Monte Carlo variance term arises. This graphic
        builds on~\cite{molnar_relating_2023}.}\label{fig:error-graph}
\end{figure}

\paragraph{Feature effect error estimation.} Next, we aim to estimate these error terms.
$\mathbb{E}_F$ and $\text{Var}_F$ can be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets, each independently sampled from the DGP. We propose the following estimators:  % chktex 13

\begin{equation}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
    \label{eq:mse-estimator}
\end{equation}
\begin{equation}
    \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
    \label{eq:bias-estimator}
\end{equation}
\begin{equation}
    \begin{split}
    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) = \\
    \frac{1}{M-1}\sum_{m=1}^M{\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)}^2
    \end{split}
    \label{eq:variance-estimator}
\end{equation}

\noindent The variance estimator is the same as in~\cite{molnar_relating_2023},
capturing both the variance in the model fits and the variance due to
MC integration. For the Monte Carlo variance of estimating the ground truth by
Monte Carlo integration, i.e.\ omitting the model, we propose the following estimator:
\begin{equation}
    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K{(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))}^2
    \label{eq:mc-variance-estimator-groundtruth}
\end{equation}

\noindent To estimate the Monte Carlo variance for feature effects of models
given in equation~\ref{eq:mse-decomposition}, we have to consider multiple
Monte Carlo iterations for multiple model refits, yielding the following estimator:

\begin{equation}
    \begin{split}
        \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
        {\left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)}^2
    \end{split}
    \label{eq:mc-variance-estimator-model}
\end{equation}

\noindent While these estimators are based on the PDP, they
can be defined analogously for ALE.  % chktex 13
% For a more convenient analysis of the errors, we suggest
% an aggregation of the error terms over the marginal distribution of $X_S$
% (e.g., by averaging over the grid points if chosen appropriately)
% to obtain a single error measure per feature effect.
It is important to note that most of these estimators are rather
impractical as they require knowledge of the ground truth. Nonetheless,
we rather use them to understand the behavior of the feature
effect error components under different dataset choices. This
distinguishes our approach from previous works such as~\cite{molnar_relating_2023},
which mostly focus on providing practical variance estimates to provide
confidence intervals for uncertainty quantification in feature effects.

@@

To address the central research question of this paper, which data to
choose for feature effect estimation, we conduct a comprehensive simulation
study. For different models,
datasets, and dataset sizes, we estimate the feature effects PDP and ALE on
training data, validation data, and in a cross-validated manner. We then
compute the feature effect error as MSE against the ground truth
feature effects, and decompose it into bias and variance components. By
comparing these error terms across the different estimation strategies
(training, validation, cross-validation), we aim to provide nuanced empirical
insights into the implications of the data choice on feature effect estimation.
To deepen our analysis, we conduct two ablation studies: (1)~decomposing feature
effect variance into model variance and Monte Carlo variance, and (2)~examining
the influence of dataset size on Monte Carlo variance.
In the following, we describe our experimental set-up for the main simulation
study (\textsc{Section~\ref{sec:experiment-main}}) and the ablation studies
(\textsc{Section~\ref{sec:experiment-ablation}}).

@@

\begin{itemize}[label=--]
    \item \textbf{SimpleNormalCorrelated} consists of four
          standard-normally distributed features, where the first two
          are strongly correlated ($\rho = 0.9$) while the others are
          independent dummy variables. The target variable is given by:
          \begin{equation}
              f_1(\boldsymbol{x}) = x_1 + \frac{x_2^2}{2} + x_1 x_2
          \end{equation}
          This test function, inspired by~\cite{liu_model_2018}, focuses
          on investigating the impact of correlation and interactions.
    \item \textbf{Friedman1} implements the classical Friedman1 benchmark
          function~\cite{breiman_bagging_1996,friedman_multivariate_1991}
          with seven mutually independent, uniformly distributed features
          between 0 and 1:
          \begin{equation}
              f_2(\boldsymbol{x}) = 10 \sin(\pi x_1 x_2) + 20{(x_3 - \frac{1}{2})}^2 + 10 x_4 + 5 x_5
          \end{equation}
          This dataset includes a mix of linear and non-linear effects with
          interactions.
    \item \textbf{Feynman I.29.16} is based on the Feynman equation I.29.16
          describing wave interference, comprising six independent features. Following
          the refined sampling strategies from Matsubara et al.~\cite{matsubara_rethinking_2024},
          we use two log-uniformly distributed variables on $[0.1, 10]$, two angles uniformly
          distributed over $[0, 2\pi]$, and two uniformly distributed dummy variables on [0,1]:
          \begin{equation}
              f_3(\boldsymbol{x}) = \sqrt{x_1^2 + x_2^2 + 2 x_1 x_2 \cos(\theta_1 - \theta_2)}
          \end{equation}
          This dataset is designed to reflect a physics-based relationship for enhanced real-world relevance.

@@

\noindent To generate the datasets, we add a standard-normally distributed noise term $\epsilon$ to each function,
scaled by a signal-to-noise ratio of five\footnote{To determine the noise scaling factor,
    we compute the standard deviation of the signal over 100,000 randomly drawn samples of $y$ and
    divide by the signal-to-noise ratio of five.}. For each test function, we consider two dataset sizes:
$(n_{train}=1000, n_{val}=250)$ and $(n_{train}=8000, n_{val}=2000)$.

@@

\paragraph{Feature effect estimation.}
For the trained models, we estimate the feature effects
$\widehat{PDP}_{\hat{f},S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
Monte Carlo integration, we use for each the full training and validation set
as well as a cross-validation strategy. In the latter case, we use a 5-fold
cross-validation on the combined samples of training and validation set, in
each fold fit the model on four folds, and estimate the feature effects on the
remaining fold. We then average the feature effects point-wise over all folds.

We also estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
$\widehat{ALE}_{f,S}$ as the complexity of the Feynman equation makes
an analytical computation infeasible. For their Monte Carlo integration, we use
10,000 samples additionally drawn from the DGP, which introduces only a
negligible additional variance as we show in (Section~\ref{sec:results-dataset-size}). We use 100 grid
points, defined by the quantiles of the theoretical distribution of $X_S$ to
enable a comparison across feature effects. Additionally, we center the curves
after estimation to have a mean effect of $0$ w.r.t.\ the grid points, and we
omit the first and the last grid point after estimation to avoid boundary
effects particularly occuring in ALE plots.

\paragraph{Feature effect errors.} To quantify the error of the estimated feature effects, we compute the MSE of
the estimated model feature effects against the estimated ground truth effects,
as well as their bias and variance. We use the
estimators defined in equations~(\ref{eq:mse-estimator}) to
(\ref{eq:variance-estimator}), using $\widehat{PDP}_{f,S}$ instead of
$PDP_{f,S}$ as ground truth (analogously for ALE). We repeat each
dataset-size-model combination $M=30$ times to estimate the error terms, where
each repetition involves drawing a new training and validation set from the
DGP, refitting the models, and estimating the feature effects. To aggregate the
errors into a single number per feature effect, we average them over the
marginal distribution of $X_S$ by averaging over the grid points.
